Indexing Variation Graphs\thanks{Supported by the Wellcome Trust grant 098051.}

### 2016-11-01 paper/paper.tex
Final ALENEX 2017 version of the paper

We compared the query performance of the \orderk{128} GCSA2 to several FM-indexes for the reference sequence. SSA is the SDSL implementation (\texttt{csa\_wt<>} with SA sample period 17) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT. As the default FM-index in SDSL, it prioritizes query performance over compression. BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with the default SA sample period 32). Optimized for DNA sequences, BWA indexes both the reference and its reverse complement.

As building the original GCSA requires around $65n$ bytes of memory for a path graph with $n$ nodes, we could not compare GCSA and GCSA2 directly on a system with 256~gigabytes of memory. Instead, we used \emph{RLCSA} \cite{Maekinen2010} (May~2016 version with SA sample period 32) as a proxy. The RLCSA is an FM-index for repetitive sequence collections using the same basic components as the original GCSA. With a mixed query load, RLCSA is 1.5x to 3x faster than GCSA, depending on algorithmic overhead and the mix of $\find$ and $\locate$ queries \cite{Siren2014}.

@@

GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6. On the average, each sample contains 1.53~values, and GCSA2 retrieves each distinct value 2.84~times, making the amount of work comparable to sample period 19.7. SSA with sample period 17 is 2.4x faster than GCSA2, highlighting the slowdown from checking multiple bitvectors $B_{c}$ in $\LF(i)$ queries, as well as from filtering out duplicates. BWA is closer to GCSA2 in $\locate$ performance, which can be explained by sparser sampling than in SSA. RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges. Assuming that $\locate$ queries in GCSA are 3x slower than RLCSA, as GCSA does use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.

>>>>

We compared the query performance of the \orderk{128} GCSA2 to several FM-indexes for the reference sequence. SSA is the SDSL implementation (\texttt{csa\_wt<>}) of the \emph{succinct suffix array} \cite{Maekinen2005}, using a Huffman-shaped wavelet tree on top of the BWT. As the default FM-index in SDSL, it prioritizes query performance over compression. We used SSA with SA sample period 17 for good $\locate$ performance. BWA is the FM-index in the \emph{Burrows-Wheeler Aligner} \cite{Li2009} (version~0.7.15 with the default SA sample period 32). Optimized for DNA sequences, BWA indexes both the reference and its reverse complement.

As building the original GCSA requires around $65n$ bytes of memory for a path graph with $n$ nodes, we could not compare GCSA and GCSA2 directly on a system with 256~gigabytes of memory. Instead, we used \emph{RLCSA} \cite{Maekinen2010} (May~2016 version) as a proxy. The RLCSA is an FM-index for repetitive sequence collections using the same basic components as the original GCSA. Under a mixed query load, RLCSA with SA sample period 32 is 1.5x to 3x faster than GCSA, depending on algorithmic overhead and the mix of $\find$ and $\locate$ queries \cite{Siren2014}.

@@

GCSA2 uses denser SA sampling than the other indexes, with effective sample period 10.6. On the average, GCSA2 calls $\locate$ for 2.84~nodes per distinct value, making the amount of work comparable to sample period 30.2. SSA with sample period 17 is 2.4x faster than GCSA2, mostly because it has to do less work. BWA with sample period 32 is closer to GCSA2 in $\locate$ performance. RLCSA is slower than the other indexes, but the difference is smaller than with $\find$ queries due to the optimizations for retrieving suffix array ranges. Assuming that $\locate$ queries are 3x slower in GCSA than in RLCSA, as GCSA does not use the optimizations, we can estimate that GCSA2 is 4x faster than GCSA.
